---
title: 'EEB319: Intro to R'
author: "M. Jarvis-Cross, Last Updated: January 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is R?
R is a computer programming language and environment for statistical computing and graphics. R is open-source, which means you can use it for free, and is maintained and updated by its users, who can write and upload "packages", which are units of sharable code that extend R's basic toolkit. As of June 2019, CRAN (Comprehensive R Archive Network), hosts over 14,000 packages, which means that if you're working on a problem, another R user likely has a solution! For example, you can use a package called ["lme4"](https://cran.r-project.org/web/packages/lme4/lme4.pdf) to fit linear or generalized linear mixed-effects models, ["popbio"](https://cran.r-project.org/web/packages/popbio/popbio.pdf) to construct and analyse projection matrix models, ["ggmap"](https://cran.r-project.org/web/packages/ggmap/ggmap.pdf) to visualize spatial data on maps, and ["wesanderson"](https://github.com/karthik/wesanderson) to make your plots look like they were taken from the set of Moonrise Kingdom!  

**Resources:**  
[R Project](https://www.r-project.org/about.html)   
[Big Book of R](https://www.bigbookofr.com/index.html)  

## Downloading R and R Studio (DO THIS BEFORE LAB 1 STARTS, PLEASE!)    
Download and install R onto your computer [here](https://utstat.toronto.edu/cran/). After downloading/installing R, you'll want to download/install [RStudio](https://www.rstudio.com/products/rstudio/), which is just R with a graphical user interface (GUI). Try opening R Studio to make sure it works. You should see four windows: the top-left window is your editor, where you can write, run, and save scripts. The bottom-left window is your console/terminal, where you can run code without saving it in a script. The top-right window is your environment, where you can find any objects you've defined. The bottom-right window is where you can navigate between files in your working directory, view plots you've generated, or read package documentation. Don't panic! We'll try to go through all of this in detail during lab.  

## R Markdown
This document, along with most of the labs in this course, and your assignment submissions, will be written using R Markdown. R Markdown is a file format for making dynamic documents (documents that can be continually edited and updated) with R. An R Markdown document is written in markdown (an easy-to-write plain text format) and contains chunks of embedded R code. You can open and edit R Markdown files (.Rmd) in R Studio. To learn more about R Markdown, go [here](https://rmarkdown.rstudio.com/articles_intro.html).   

**AN IMPORTANT NOTE:** When we write Markdown documents, we can export .Rmd files to PDF, Word, or HTML via a process called "knitting". You should be able to knit this file ("EEB319IntroR.Rmd") to HTML by clicking "knit" at the top left of this page. Once R has finished knitting this file to HTML, a preview window will pop up. You can view the knitted document in HTML by clicking "Open in Browser".  

**ANOTHER IMPORTANT NOTE:** As previously mentioned, Markdown documents are neat because they can integrate chunks of text with chunks of code. To learn how we integrate a chunk of code, check of the example we use to set a working directory, below. Take a look at the "eval" argument. If we set **eval = TRUE**, R runs that chunk of code when knitting a document. If we set **eval = FALSE**, R will include the chunk in the knitted document, but won't run the code. In the "setwd()" example below, we've set eval = FALSE because we have different computers, and as such, the line we've written won't run on your computer. In upcoming labs, you'll notice that we've set eval = FALSE in some of our examples, and have included an empty code chunk, where we're set eval = TRUE, for you to write your own code below.  

## Let's Get Started w/ an Example!  
### Setting Your Working Directory:  
We use working directories to tell R where we want to look for and store files. We can start by making a folder called "R_Projects", and a subfolder called "EEB319" on our desktops. If you right-click on that subfolder, you should be able to see its pathname (e.g. /Users/mjarviscross/Desktop/R_Projects/EEB319). As a side-note, your working directory can be thought of similarly to how you'd think about any folder. For example, let's download the course files from Quercus into our "EEB319" folders, and create subfolders for different labs.  

There are two ways to set your working directory: 
  
#### setwd():  
Use **setwd()** to set your working directory by including the pathname of your EEB319 folder:  
```{r, eval=FALSE}
setwd("/Users/mjarviscross/Desktop/R_Projects/EEB319")
```
This is the easier option, but not the best...  

#### Start a Project:  
If you set your working directory using **setwd()**, you'll be able to save all your files in the same place, but won't be able to save any of your objects or outputs. If you set your working directory by [creating an R Project](https://r4ds.had.co.nz/workflow-projects.html), you'll be able to save all your files in the same place *and* save all your objects and outputs. This will save you a lot of time in the long-run, and make your work reproducible.  

Go to the top-right corner of your R Studio window and click "Project: (None)". Then, click the first option in the drop-down menu, "New Project". A new window will pop-up, and give you three options: "New Directory", "Existing Directory", and "Version Control". Click "Existing Directory", and use the browse button to navigate to your "EEB319" folder. Finally, click "Create Project" in the bottom-right of the pop-up window, and you're ready to go!  

### An Example for Today's Lab: Within-Host Disease Dynamics  
To get more comfortable with R, let's go through an example! In their 1994 paper, "[Within-host population dynamics and the evolution and maintenance of microparasite virulence](https://www.journals.uchicago.edu/doi/pdfplus/10.1086/285686?casa_token=9YROqojZVtEAAAAA:5uN5aPqif9X6Ra98p2L0Dh-zsebHg855MPUw85obnC15626w55jhNWbFE_dorGoqWMhFjiiFhK3B)", Antia et al. presented a mathematical model describing the within-host population dynamics of an invading microparasite and its host's immune response.  

\begin{align}
    &\frac{dP}{dt} = rP - kPI \text{,  if P < D and P ➝ 0 if P > D}&\\
    &\frac{dI}{dt} = ρI\frac{P}{P + ϕ}&\\
\end{align}  

Where:  
```{r, echo=FALSE, results='asis'}
Parameters <- c("r, Intrinsic growth rate of parasite", "k, Rate of destruction of parasite by host immune system", "ρ, Maximum growth rate of the immune system", "𝜙, Parasite density at which growth rate of host immune system is half its maximum","D, Lethal within-host parasite density")
Values <- c("0.1-2.0", "0.001", "1", "1000", "10<sup>9</sup>")
parms_table <- data.frame(Parameters, Values)
library(knitr)
kable(parms_table, caption = "Table 1")
```

This is an ODE (ordinary differential equation) model, and is comprised of a series of differential equations. The equations in this model represent the parasite (P) and host immune cell (I) populations through time. The first term in the first equation denotes the intrinsic growth of the parasite population, while the second term denotes the destruction of the parasite by the host's immune system. This model assumes that the rate of proliferation of the immune cell population is proportional to the size of the parasite population when the parasite population is small, and saturates when the parasite population grows sufficiently large. Try thinking about how the second equation denotes this relationship mathematically.  
  
<p style="color: red;">**Question: Physiologically, why might the authors make this assumption?**</p>

For today's lab, we'll be simulating a version of this model forward in time. Antia et al.'s original model is written in *continuous time*, but we'll be using a version of this model written in *discrete time*.  

\begin{align}
    &P_{t+1} = (1 + r⋅ts - (1-e^{-k⋅I_t⋅ts}))P_t&\\
    &I_{t+1} = (1 + ρ\frac{P}{P + ϕ}ts)I_t&\\
\end{align}  

<p style="color: red;">**Question: Why did we exponentiate the second term in the first equation?**</p>  

Now, to simulate this model forward in time, open the Lab 1 Excel sheet; come back here when you've finished simulating your data!  

**...**  

Welcome back! The Excel sheet asked you to save your simulated data as a .csv. Let's bring those data into R!  

### Importing Data  
Remember to make sure your data are stored in a .csv file, in your *working directory*. **You can import your data using read.csv():**  

Make sure to give your data set a name so you can call on it later; here we're calling our data set "InfecData". We use "**<-**" ("option" + "-")  as an assignment operator: it assigns a value to an object. Here, the value is our data set, and the object is "InfecData".  
```{r, eval=TRUE}
# NOTE 1: We can put a "#" at the beginning of a line of code if we don't want R to run that line. These lines are called "comments" and can be used to leave yourself notes explaining what your code does, why you chose to do something a certain way, etc.!

# NOTE 2: To run a line or chunk of code in your editor, use "command + Enter" on Mac, and "Ctrl + Enter" on Windows. If you're working in your console, you can just use "Enter".

InfecData <- read.csv("ModelSimulation.csv") # NOTE: that you need to include the entire pathname for your .csv within the quotation marks.
```

In R, we can store different variables and objects in different ways. It's important to know the "class" of a variable or object because depending on its class, we'll have to treat the variable/object differently. We can see what kind of object our data are stored in using **class()**:  
```{r, eval=TRUE}
class(InfecData)
```
Our data are stored in a data.frame! A data frame is the most common way of storing data, and is the data structure we most often use when analysing or visualizing data. Technically, data frames are just lists of equal-length vectors, making them easy to manipulate and play around in.  

### Exploring Data  
**Let's take a look at our data!** You can view the data set in its entirety by using **View()**:  
```{r, eval=TRUE}
View(InfecData)
```

Do your data look similar to how they looked in Excel? You should have 501 rows and 3 columns (time, parasite abundance, and host immune cell abundace). Next, let's explore our data!  

#### Using Functions  
**You can explore these data in a few ways using functions!** In R, functions are commands you use to tell R to perform some task. The good news is, you've already used a few functions ("read.csv()", "class()", "View()"). To use functions generally, you type out the function's name and open brackets, "()". Inside your brackets, you'll include your arguments, which tell R where and how you're using a function. R has a lot of built-in functions, but we can find new functions in packages, and even write our own (but we'll talk more about that later).  
```{r, eval=TRUE}
# If you're not sure what a function does or how to make it work, you can ask R by using "?":
?head # The help tab in your bottom-right window will tell you what "head()" does and how to use it!

# We can see the first/last few lines of our data set using head() and tail():
head(InfecData)
tail(InfecData)

# And we can specify how many lines we'd like to see by adding an argument for "n":
head(InfecData, n = 3)

# We can summarize our data using summary():
summary(InfecData)

# We can apply these same functions to a sebset of our data set using "$" to specify the column we're interested in:
max(InfecData$Parasites) # Shows us the maximum abundance of parasites present in the host at a given time
which(InfecData$Parasites == max(InfecData$Parasites)) # Reads: Where in the "Parasites" column of the the data set is the data point equal to (==) the maximum abundance of parasites (shows us when the parasite population peaks).

# To use a function from a package, we first have to install and load the package:
# install.packages("cowsay") # Installing
library(cowsay) # Loading
?cowsay # Exploring the package
?say # Exploring the function
say("This is one of R's many user-generated useless-but-fun packages!") # Using the package's function
say("Maddie's writing this in October, so I'm a ghost!", by = "ghost") # Adding an argument to change how we're using the function
```
  
### Plotting Data  
**We can plot our data using plot().** We need to give plot() the data we want to use for the x-axis first, followed by the data we want to use for the y-axis. Here, we want "time" on the x-axis, and the abundaces of parasites and immunce cells on the y-axis. In the example below, the first line plots the population abundances of the parasite through time, and the second line adds the population abundances of the immune cells through time to the plot. The third line adds a legend!  

We can use a few additional arguments to make aesthetic changes to our plot:  

* **type** changes the line type  
* **col** changes the colours of our data  
* **lwd** changes the line width  
* **xlab** and **ylab** change the names of our axes  

```{r, eval=TRUE}
plot(InfecData$time, InfecData$parasites, type = "l", col = "goldenrod1", lwd = 4, xlab="Time", ylab="Abundance")
lines(InfecData$time, InfecData$immune_cells, col = "cornflowerblue", lwd = 4)
legend("topleft", legend = c("Parasites", "Immune Cells"), col = c("goldenrod1", "cornflowerblue"), lty = 1, lwd  =4)
```

Now, try to make a plot like the one above, with some aesthetic changes! You can use [STHDA](http://www.sthda.com/english/wiki/line-plots-r-base-graphs) and [Towards Data Science](https://towardsdatascience.com/base-plotting-in-r-eb365da06b22) as references!  
```{r, eval=TRUE}
# YOUR WORK HERE!
```

<p style="color: red;">**Question: How might we interpret these data? Do these data seem to represent a chronic infection? An infection that can be cleared? An infection that ends with the death of the host? Explain.**</p>

## After Getting Started...  
### For Loops  
For loops are used to repeat a chunk of code some number of times. For loops are easy to write once you get the hang of them, and can be used to save you a lot of time! The general syntax for a for loop is as follows:  
```{r, eval=FALSE}
for (i in sequence){
  perform some operation 
}
```
Here, "i" is what we call the "iterator", and "perform some operation" is the "iterable".  
  
**Example:** 
```{r, eval=TRUE}
vec <- c() # Makes an empty vector to store the outputs of our for loop in
for (i in 1:10){
  vec[i] <- i + 1 # The value at position i in our vector is i + 1
}
# Means that the value at position 1 in our vector is 1 + 1, the value at position 2 in our vectos is 2 + 1, etc.

print(vec) # To show our resulting vector
```

### if... else Statements  
if... else statements are conditional statements that are used to evaluate an input, and perform different tasks based on if the input passes some test. The general syntax for an if... else statement is as follows:  
```{r, eval=FALSE}
if (test_expression){
  statement
}

if (test_expression){
  statement
} else{
  statement2
}
```
**Example:**  
```{r, eval=TRUE}
x <- 7 # Assign x some value
if (x > 2){
  print("It's a bones day")
}
```
**And another one...**  
```{r, eval=TRUE}
if (x == 5){
  print("It's a bones day!")
} else {
  print("It's a no bones day!")
}
```

### Combining For Loops and if... else Statements  
For loops and if... else statements are often used together by nesting the if... else statement within the for loop.    
``` {r, eval=TRUE}
vec2 <- c(seq(1:10)) # Makes a vector with values 1-10
vec3 <- c() # Another empty vector to store the outputs of our for loop in 
for (i in vec2){
  if (vec2[i] <= 5){
    vec3[i] <- "It's a bones day!"
  } else {
    vec3[i] <- "It's a no bones day!"
  }
}
print(vec3)
```
You can also nest for loops inside of other for loops and so on. BUT! A few layers of for loops can really slow down your computation. When that happens, we're better off writing a function. In R, functions are self-contained modules of code that accomplish a specific task. To use a function, you give its required arguments, the function completes its task, and gives you back an output. The general syntax for a function is as follows:  
```{r, eval=FALSE}
function_name <- function(argument){
  statement
}
```
**Example:**  
```{r, eval=TRUE}
Eqn_Func <- function(x){
  y <- 2*x + 4
  return(y)
}

# The function needs us to give a value of x. If we say x = 2, then the function calculates y according to the statment we wrote, y = 2x + 4
Eqn_Func(2)
```
We can also put a function inside a for loop (and so on, and so on...)  
**NOTE:** There's a lot going on in this example! It's okay if this is a bit confusing!
```{r, eval=TRUE}
# This function takes initial parasite and immune cell population abundances, and, according to a model we've written, determines the population abundances at the next time step
Sim_Func <- function(Para_Last, Immune_Last){
  # Parameter values
  ts = 0.1
  r = 0.3; k = 0.001; p = 1; o = 1000
  
  # A discrete model of within-host infectious disease dynamics
  Para_Next = (1 + r*ts - (1-exp(-k*Immune_Last*ts)))*Para_Last
  Immune_Next = (1 + p*(Para_Last/(Para_Last + o))*ts)*Immune_Last
  
  Para_Last <- Para_Next
  Immune_Last <- Immune_Next
  
  return(c(Para_Last, Immune_Last))
}

# To project this model into the future, we need to implement the function within a for loop that can feed the function the return values from the previous iteration of the loop. 
Abundances <- data.frame() # Empty data frame to hold our outputs
Para_Last <- 1; Immune_Last <- 1 # Initial population abundances
for (i in 1:500){ # Loop will run the function 500 times, but with new "initial" abundances every time to simulate the populations forwards in time
  Output = Sim_Func(Para_Last, Immune_Last)
  Para_Last = Output[1]
  Immune_Last = Output[2]
  Addition <- c(Para_Last, Immune_Last)
  Abundances <- rbind(Abundances, Addition)
  colnames(Abundances) <- c("Parasites", "Immune_Cells")
}

# Let's see if that worked:
head(Abundances)

# Let's plot our output!
Inits <- data.frame("1", "1")
colnames(Inits) <- c("Parasites", "Immune_Cells")
Abundances <- data.frame(rbind(Inits, Abundances))
Time <- seq(0,50,0.1)
df <- cbind(Time, Abundances)
plot(df$Time, df$Parasites, type = "l", col = "goldenrod1", lwd = 4, xlab="Time", ylab="Abundance")
lines(df$Time, df$Immune_Cells, col = "cornflowerblue", lwd = 4)
legend("topleft", legend=c("Parasites", "Immune Cells"),
       col=c("goldenrod1", "cornflowerblue"), lty = 1, lwd = 4)
```

## Making Your Own Markdown Document  
For this course, you'll be submitting your lab assignments as Markdown documents. You can make your own Markdown document by clicking **File > New File > R Markdown**. After you've opened that blank Markdown document, you can start typing, using this document, and the following resources as a guide.  

* [Markdown Example](https://www.math.mcgill.ca/yyang/regression/RMarkdown/example.html)  
* [Formatting Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)  

## Learning Resources:  
* [Introduction to R by DataCamp](https://app.datacamp.com/learn/courses/free-introduction-to-r?utm_source=adwords_ppc&utm_campaignid=897699256&utm_adgroupid=47591961031&utm_device=c&utm_keyword=r%20course&utm_matchtype=p&utm_network=g&utm_adpostion=&utm_creative=262625967799&utm_targetid=kwd-301266522201&utm_loc_interest_ms=&utm_loc_physical_ms=9000970&gclid=CjwKCAjwoP6LBhBlEiwAvCcthH6H9DeWkdFYviWhblTYEdQKQpI6_JqaBcv5Q-M8RBnmAFetA1xSZRoCh9kQAvD_BwE)  
  + A free interactive course with modules to teach you the basics, vectors, matricies, factors, data frames, and lists  
  + 4 hour time committment  
* [Tips and Tricks in R Studio and R Markdown](https://towardsdatascience.com/tips-and-tricks-in-rstudio-and-r-markdown-7a464508b0c)  
  + A really great article from Towards Data Science (generally a fantastic resource!) with some tips on how to write R scripts more efficiently  
* [Stack Overflow](https://stackoverflow.com)  
  + A Q&A website for asking and answering coding questions  
  + If you're having trouble figuring out how to do something, or can't figure out why your code isn't running, check Stack Overflow! It's likely that someone's already asked and answered your question!  
  
## Marking:  

* Code with clear and concise commenting: 40%  
* Does your code run? 10%  
* Plotting: 25%  
* Interpretation questions: 25%  

   



