---
title: "321-Lab6-RMD"
author: "Leila Forsyth"
date: "July 30, 2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "hide",error = FALSE,warning = FALSE,message = FALSE)
```


## Setting up R
You need to first paste the needed .csv files, and the DO NOT OPEN r script onto your desktop

Next, you will set your working directory using the following code:

> setwd("C:/Users/UTorID/Desktop") 

```{r, set working directory, include=FALSE,echo=FALSE}
setwd("C:/Users/abuga/Dropbox/EEB321 Winter 2018 Shared Folder/2020- RMD Course development/Lab 6 Neutral Theory -check Emma") 
source('eeb321 neutral lab functions - DO NOT OPEN.r')
library(vegan)
```


Now we will set up and load the functions that we will run, using the DO NOT OPEN r script.
```{r}
source('eeb321 neutral lab functions - DO NOT OPEN.r')
```

Now we will install the 'vegan' package using the code:
```{r}
install.packages("vegan", dependencies=T,repos="http://R-Forge.R-project.org")
```

If this does not work, run the code:
```{r}
install.packages("vegan")
```

Once installed, run the code:
```{r}
library(vegan)
```


Note: you only need to install packages to your computer ONCE, afterwards, you can always call on the package using the library() or require() functions.

Now we will run an example graph for you to get familiar with the parameters.

Enter the code:
```{r}
bh.display(aii=1,aij=1,li=11,aji=1,ajj=1,lj=11,ni=5,nj=5,years=10)
```

aii= intraspecific competition effect for species i
aij= interspecific competition effect for species i on j
li= lambda for species i
aji= intraspecific competition effect for species j
ajj= interspecific competition effect for species j on i
lj= lambda for species j
ni= starting abundance of species i
nj= starting abundance of species j
years= number of years to run the simulation

Play around with the parameters to understand how competition and populaiton growth rates (lambda) influence the coexistence dynamics.

## Example figure:
```{r, example figure,fig.show=TRUE}
n1<-bh.display(aii=1,aij=1,li=11,aji=1,ajj=1,lj=11,ni=5,nj=5,years=10)
n1
```


To get the beta diversity and extinction results when each species starts at the same abundance, run the code:

```{r}
bh(aii=1,aij=1,li=11,aji=1,ajj=1,lj=11,ni=5,nj=5,years=10) 
```


## EXERCISE 1: Neutral 1

Modify this code to answer the questions in exercise 1:

```{r}
bh(aii=1,aij=1,li=11,aji=1,ajj=1,lj=11,ni=5,nj=5,years=10) 
```

## EXERCISE 2: Neutral 2

Modify this code to answer the questions in exercise 2:

```{r}
bh(aii=1,aij=1,li=21,aji=1,ajj=1,lj=21,ni=10,nj=10,years=20)
```


## EXERCISE 3: One winner

Modify this code to answer the questions in exercise 3:

```{r}
bh(aii=1,aij=1,li=7,aji=1,ajj=1,lj=10,ni=1,nj=9,years=20)
```

## EXERCISE 4: Coexistence 

Modify this code to answer the questions in exercise 4:

```{r}
bh(aii=1,aij=1,li=15,aji=1,ajj=1,lj=15,ni=1,nj=14,years=10)
```



